---
title: "Tradeseq"
author: "Claire Man"
date: "3/18/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
- For every gene in the datast, tradeSeq fits a generalized additive model (GAM) by building on the mgcv R package.  
- It then allows statistical inference on the GAM by assessing contrasts of the parameters of the fitted GAM model, ading in interpreting complex datasets.  
  
  
# Load Data  
```{r}
library(tradeSeq)
library(RColorBrewer)
library(SingleCellExperiment)
library(slingshot)

# For reproducibility
RNGversion("3.5.0")
palette(brewer.pal(8, "Dark2"))
data(countMatrix, package = "tradeSeq")
counts <- countMatrix
rm(countMatrix)
data(crv, package = "tradeSeq")
data(celltype, package = "tradeSeq")
```
  
  
# Fit trajectories using slingshot   
- use the reduced space calculated with the UMAP method.  
- cluster the data using k-menas with 7 clusters.
  
```{r}
set.seed(200)
rd <- reducedDims(crv)
cl <- kmeans(rd, centers = 7)$cluster
plot(rd, col = brewer.pal(9, "Set1")[cl], pch = 16, asp = 1,
     cex = 2/3)
```
  
- get the trajectory
```{r}
lin <- getLineages(rd, clusterLabels = cl, start.clus = 1)
crv <- getCurves(lin)
plotGeneCount(curve = crv, clusters = cl)

# colored by cell type
plotGeneCount(curve = crv, clusters = celltype, 
              title = "Colored by cell type")
```
  
  
# Determining the number of knots (?)
- The core function from tradeSeq, fitGAM, will use cubic splines as basis functions, and it tries to ensure that every lineage will end at a knot point of a smoother.  
- The number of knots should be selected to reach an optimal bias-variance trade-off for the smoother (AIC).   
```{r}
library(Matrix)
icMat <- evaluateK(counts = counts, sds = crv, k=3:20, nGenes = 200,
                   verbose=FALSE)
# alternatively:
icMat <- evaluateK(counts = counts, k=3:20, nGenes = 200,
                   pseudotime = slingPseudotime(crv, na = FALSE),
                   cellWeights = slingCurveWeights(crv))
```
  
  
# Fit Additive models  
- By default, the GAM model estimates one smoother for every lineage using the negative binomial distribution.   
- tradeSeq allows input from any trajectory inference method with the pseudotime and cellWeights arguments.
- Because cells are assigned to a lineage based on their weigths, the result of fitGAM is stochastic. 

```{r}
library(BiocParallel)
library(magrittr)
# Register BiocParallel Serial Execution (no parallelization in that case)
BiocParallel::register(BiocParallel::SerialParam())

counts <- as.matrix(counts)
sce <- fitGAM(counts = counts,
                  sds = crv)
# This takes about 1mn to run
```
  
- Can also plot the cells in reduced dimension to see where the knots are located.  
```{r}
plotGeneCount(curve = crv, counts = counts, clusters = cl,
              models = sce)
```
  
  
# Within-lineage comparisons  
## Association of gene expression with pseudotime  
- checking whether gene expression is associated with a particular lineage
- testing the null hypothesis that all smoother coefficients are equal to each other
- testing whether the smoothed gene expression is significantly changing along pseudotime
  
```{r}
assoRes <- associationTest(sce)
head(assoRes)
```
  
## Discovering progenitor marker genes
- assessing differential expression between the progenitor cell population (i.e., the starting point of a lineage) with the differentiated cell type population (i.e., the end point of a lineage).  
  
```{r}
startRes <- startVsEndTest(sce)

# Visualize the estimated smoothers for the most significant gene.
oStart <- order(startRes$waldStat, decreasing = TRUE)
sigGeneStart <- names(sce)[oStart[1]]
plotSmoothers(sce, counts, gene = sigGeneStart)
```
  
```{r}
# color the cells in UMAP space with that gene’s expression
plotGeneCount(crv, counts, gene = sigGeneStart)
```
  
## Comparing specific pseudotime values within a lineage  
- The startVsEndTest compares two points on a lineage, and by default it is comparing the inception point with the end point.  
  
```{r}
customRes <- startVsEndTest(sce, pseudotimeValues = c(0.1, 0.8))
```
  
  
# Between-lineage comparisons 
## Discovering differentiated cell type markers  
- tradeSeq can discover marker genes for the differentiated cell types by comparing the end points of the lineage-specific smoothers.
- By default, diffEndTest performs an omnibus test, testing the null hypothesis that the endpoint expression is equal for all lineages using a multivariate Wald test.  
  
```{r}
endRes <- diffEndTest(sce)

# plot the most significant gene using the plotSmoothers function.
o <- order(endRes$waldStat, decreasing = TRUE)
sigGene <- names(sce)[o[1]]
plotSmoothers(sce, counts, sigGene)
```
  
```{r}
# color the cells in UMAP space with that gene's expression
plotGeneCount(crv, counts, gene = sigGene)
```
  
## Discovering genes with different expression patterns (?)  
- The function patternTest implements a statistical method that checks whether the smoothed gene expression is equal along pseudotime between two or multiple lineages.  
  
```{r}
patternRes <- patternTest(sce)
oPat <- order(patternRes$waldStat, decreasing = TRUE)
head(rownames(patternRes)[oPat])
```
  
```{r}
plotSmoothers(sce, counts, gene = rownames(patternRes)[oPat][1])
plotGeneCount(crv, counts, gene = rownames(patternRes)[oPat][1])
```
  
- sort the genes according to the sum of square of their rank in increasing Wald statistics for the patternTest and their rank in decreasing Wald statistics for the diffEndTest.  
  
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
```
  
```{r}
compare <- inner_join(patternRes %>% mutate(Gene = rownames(patternRes),
                                            pattern = waldStat) %>%
                                     select(Gene, pattern),
                      endRes %>% mutate(Gene = rownames(endRes),
                                        end = waldStat) %>%
                                 select(Gene, end),
                      by = c("Gene" = "Gene")) %>%
           mutate(transientScore = (min_rank(desc(end)))^2 +
                                   (dense_rank(pattern))^2)

ggplot(compare, aes(x = log(pattern), y = log(end))) +
  geom_point(aes(col = transientScore)) +
  labs(x = "patternTest Wald Statistic (log scale)",
       y = "diffEndTest Wald Statistic (log scale)") +
  scale_color_continuous(low = "yellow", high = "red") +
  theme_classic()
```
  
```{r}
# visualize the expression in UMAP space of the top gene.  

topTransient <- (compare %>% arrange(desc(transientScore)))[1, "Gene"]
plotSmoothers(sce, counts, gene = topTransient)

plotGeneCount(crv, counts, gene = topTransient)
```
  
## Early Drivers of differentiation  
- find a list of genes that are differentially expressed around the separation of two or multiple lineages.  
- The function earlyDETest implements a statistical method to tests the null hypothesis of whether the smoothers are equal between two user-specified knots by building on the patternTest, but restricting itself to a particular location of the smoothers.  
  
```{r}
plotGeneCount(curve = crv, counts = counts, clusters = cl,
              models = sce)
```
  
```{r}
earlyDERes <- earlyDETest(sce, knots = c(1, 2))
oEarly <- order(earlyDERes$waldStat, decreasing = TRUE)
head(rownames(earlyDERes)[oEarly])
```
  
```{r}
plotSmoothers(sce, counts, gene = rownames(earlyDERes)[oEarly][2])
```
  
```{r}
plotGeneCount(crv, counts, gene = rownames(earlyDERes)[oEarly][2])
```
  
  
# Clustering of genes according to their expression pattern  
- tradeSeq provides the functionality to cluster genes according to their expression pattern along the lineages with the clusterExpressionPatterns function  
  
```{r}
library(clusterExperiment)
```
  
```{r}
nPointsClus <- 20
clusPat <- clusterExpressionPatterns(sce, nPoints = nPointsClus,
                                     genes = rownames(counts)[1:200])

clusterLabels <- primaryCluster(clusPat$rsec)
```
  
```{r}
library(cowplot)
```
  
```{r}
cUniq <- unique(clusterLabels)
cUniq <- cUniq[!cUniq == -1] # remove unclustered genes

plots <- list()
for (xx in cUniq[1:4]) {
  cId <- which(clusterLabels == xx)
  p <- ggplot(data = data.frame(x = 1:nPointsClus,
                                y = rep(range(clusPat$yhatScaled[cId, ]),
                                        nPointsClus / 2)),
              aes(x = x, y = y)) +
    geom_point(alpha = 0) +
    labs(title = paste0("Cluster ", xx),  x = "Pseudotime", y = "Normalized expression") +
    theme_classic()
  for (ii in 1:length(cId)) {
    geneId <- rownames(clusPat$yhatScaled)[cId[ii]]
    p <- p +
      geom_line(data = data.frame(x = rep(1:nPointsClus, 2),
                                  y = clusPat$yhatScaled[geneId, ],
                                  lineage = rep(0:1, each = nPointsClus)),
                aes(col = as.character(lineage), group = lineage), lwd = 1.5)
  }
  p <- p + guides(color = FALSE) +
    scale_color_manual(values = c("orange", "darkseagreen3"),
                       breaks = c("0", "1"))  
  plots[[as.character(xx)]] <- p
}
plots$ncol <- 2
do.call(plot_grid, plots)
```


